\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage[hidelinks]{hyperref}
\usepackage[margin=2.5cm]{geometry}
\usetikzlibrary{arrows.meta,calc,positioning,shapes.geometric}

\title{Automatic Image Analysis Script}
\author{}
\date{\today}

\begin{document}

\maketitle
\tableofcontents
\newpage

\input{intro}

\section{Core Topics}

The following sections collect the technical building blocks that will be
expanded throughout the script. Each topic should connect back to the central
question from the introduction: what can be inferred from image measurements,
and what assumptions make that inference possible?

\subsection{Image Formation and Representation}

Add notes on image acquisition, sampling, color spaces, filtering, and basic
image transformations.

A pinhole camera model is often the first geometric abstraction:
\begin{equation}
  \lambda
  \begin{bmatrix}
    x \\ y \\ 1
  \end{bmatrix}
  =
  K
  \begin{bmatrix}
    R & t
  \end{bmatrix}
  \begin{bmatrix}
    X \\ Y \\ Z \\ 1
  \end{bmatrix}.
\end{equation}
\noindent
It states that a 3D point $(X,Y,Z)$ is projected to image coordinates $(x,y)$
through camera intrinsics $K$ and pose $(R,t)$, up to an unknown scale
$\lambda$.

\subsection{Learning-Based Methods}

Add notes on neural networks, model training, evaluation, and uncertainty.

For supervised recognition, the training set supplies pairs $(I_i,y_i)$ and the
model parameters $\theta$ are chosen by minimizing an empirical loss:
\begin{equation}
  \hat{\theta} = \arg\min_\theta
  \frac{1}{N}\sum_{i=1}^{N}\ell\!\left(g_\theta(I_i), y_i\right).
\end{equation}
\noindent
This shifts part of the modeling burden from hand-designed rules to data,
architecture, and evaluation.

\subsection{Visual Object Models}

Add notes on feature representations, object models, detection, and recognition.

\subsection{Probabilistic Estimation}

Add notes on Bayesian reasoning, probabilistic models, and parameter estimation.

\section{Exercises}

Use this section for worked exercise notes and derivations.

\end{document}
