\section{Neural Networks and Backpropagation}

Neural networks are one way to replace manually designed decision rules by
learned functions. For image analysis, the input may be an image, a patch, or a
feature vector, and the output may be a class score, a probability, a
segmentation label, or another task-specific prediction. The central idea is
simple: compose many parameterized transformations and fit their parameters
from examples.

Szeliski's treatment of deep learning, especially the part on deep neural
networks and backpropagation, is a useful companion for this topic. The focus
here is narrower: understand the computational structure of a feed-forward
network and why backpropagation is the efficient way to compute the gradients
needed for training.

\subsection{From Perceptrons to Multilayer Networks}

A perceptron is a linear classifier. Given an input vector $x$, weights $w$,
and bias $b$, it predicts a binary label by thresholding an affine score:
\begin{equation}
  f(x) =
  \begin{cases}
    1, & w^\top x + b > 0,\\
    -1, & \text{otherwise}.
  \end{cases}
\end{equation}
\noindent
The decision boundary is the hyperplane $w^\top x + b = 0$. This makes the
perceptron easy to interpret, but also limited: if the classes cannot be
separated by a linear boundary, the standard perceptron cannot solve the
problem. The XOR problem is the usual minimal example.

During perceptron training, an incorrectly predicted example changes the
weights in the direction of the input:
\begin{equation}
  w_i^{t+1} = w_i^t + \eta\left(y - f(x)\right)x_i ,
\end{equation}
\noindent
where $\eta$ is the learning rate. The rule is useful historically because it
already shows the pattern of learning from prediction error, but modern neural
networks need a smoother and more general gradient-based view.

A multilayer perceptron (MLP) overcomes the purely linear limitation by stacking
several affine maps and nonlinear activation functions. For layer $l$,
\begin{equation}
  h^l = W^l x^l + b^l,\qquad
  y^l = \phi(h^l),\qquad
  x^{l+1}=y^l .
\end{equation}
\noindent
Without the nonlinearity $\phi$, stacking layers would still collapse into one
linear transformation. The nonlinearities make it possible to build complex
decision boundaries from many simple units.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
      node distance=1.15cm,
      unit/.style={circle, draw, minimum size=0.52cm, inner sep=0pt},
      label/.style={align=center, font=\small},
      arrow/.style={-{Latex[length=2.2mm]}, thick}
    ]
    \node[unit, fill=green!8] (x1) {};
    \node[unit, fill=green!8, below=0.55cm of x1] (x2) {};
    \node[unit, fill=green!8, below=0.55cm of x2] (x3) {};

    \node[unit, fill=blue!8, right=1.5cm of x1, yshift=-0.28cm] (h11) {};
    \node[unit, fill=blue!8, below=0.55cm of h11] (h12) {};
    \node[unit, fill=blue!8, below=0.55cm of h12] (h13) {};

    \node[unit, fill=blue!8, right=1.5cm of h11] (h21) {};
    \node[unit, fill=blue!8, below=0.55cm of h21] (h22) {};
    \node[unit, fill=blue!8, below=0.55cm of h22] (h23) {};

    \node[unit, fill=orange!12, right=1.5cm of h22] (out) {};

    \foreach \a in {x1,x2,x3} {
      \foreach \b in {h11,h12,h13} {
        \draw[arrow, gray!70] (\a) -- (\b);
      }
    }
    \foreach \a in {h11,h12,h13} {
      \foreach \b in {h21,h22,h23} {
        \draw[arrow, gray!70] (\a) -- (\b);
      }
    }
    \foreach \a in {h21,h22,h23} {
      \draw[arrow, gray!70] (\a) -- (out);
    }

    \node[label, above=0.28cm of x1] {input};
    \node[label, above=0.28cm of h11] {hidden};
    \node[label, above=0.28cm of h21] {hidden};
    \node[label, above=0.28cm of out] {output};
  \end{tikzpicture}
  \caption{A feed-forward network composes affine transformations and
  nonlinearities. Training changes the weights on the connections.}
\end{figure}

\noindent
Common activation functions include the logistic sigmoid, hyperbolic tangent,
and rectified linear unit:
\begin{align}
  \sigma(z) &= \frac{1}{1 + e^{-z}},
  & \sigma'(z) &= \sigma(z)\left(1-\sigma(z)\right),\\
  \tanh'(z) &= 1-\tanh^2(z),
  & \operatorname{ReLU}(z) &= \max(0,z).
\end{align}
\noindent
For ReLU, the derivative is $1$ for positive inputs and $0$ for negative
inputs; at zero one chooses a convenient subgradient in implementations. This
simple derivative is one reason ReLU-style activations are computationally
attractive.

\subsection{Losses, Optimization, and the Need for Gradients}

Training starts with examples $(x_i,y_i)$ and a network $f_\theta$ with
parameters $\theta$, meaning all weights and biases. A loss function measures
how bad a prediction is. Over a dataset, the empirical objective is
\begin{equation}
  L(\theta) =
  \sum_{i=1}^{N} \ell\!\left(f_\theta(x_i), y_i\right).
\end{equation}
\noindent
For regression, $\ell$ may be a squared error. For classification, it is often
cross-entropy applied to class scores or probabilities. The exact loss depends
on the task, but the training problem has the same form: find parameters that
make the loss small.

Stochastic gradient descent updates parameters using a small batch
$\mathcal{B}_t$ instead of the full dataset:
\begin{equation}
  \theta_{t+1}
  =
  \theta_t
  -
  \alpha
  \sum_{i\in\mathcal{B}_t}
  \frac{\partial \ell_i(\theta_t)}{\partial \theta},
\end{equation}
\noindent
where $\alpha$ is the learning rate. The practical difficulty is not writing
this update down; it is computing the partial derivatives for millions of
parameters quickly and without duplicating work. Backpropagation solves exactly
that problem.

\begin{table}[ht]
  \centering
  \begin{tabular}{@{}p{3.0cm}p{5.1cm}p{5.1cm}@{}}
    \toprule
    Component & Role in the network & Training question \\
    \midrule
    Weights and biases & Store the adjustable model parameters &
      How should each parameter change to reduce the loss? \\
    Activations & Introduce nonlinear transformations between layers &
      Which units are active and how do they transmit gradients? \\
    Loss function & Converts predictions and targets into one scalar &
      What does ``better'' mean for this task? \\
    Optimizer & Turns gradients into parameter updates &
      How large and in which direction should the next step be? \\
    \bottomrule
  \end{tabular}
  \caption{Training a neural network requires both a model architecture and a
  procedure for turning prediction errors into parameter updates.}
\end{table}

\subsection{Backpropagation}

Backpropagation is an efficient application of the chain rule to a computational
graph. The forward pass computes the network output and stores intermediate
quantities. The backward pass starts from the derivative of the loss with
respect to the output and propagates sensitivity information back through the
same operations in reverse order.

It is helpful to name the quantities at one layer before writing the gradient.
Let $h_k$ be the activation vector that enters layer $k$. The trainable
parameters of this layer are a weight matrix $\Omega_k$ and a bias vector
$\beta_k$. The affine part of the layer first forms the pre-activation vector
\begin{equation}
  f_k = \beta_k + \Omega_k h_k,
\end{equation}
\noindent
and a nonlinearity would then produce the next activation, for example
$h_{k+1}=\phi(f_k)$. With this notation, the gradient with respect to the weight
matrix has the structure
\begin{equation}
  \frac{\partial \ell_i}{\partial \Omega_k}
  =
  \frac{\partial \ell_i}{\partial f_k}\,h_k^\top .
\end{equation}
\noindent
This formula captures a useful intuition from the lecture: the effect of a
weight depends both on the upstream error signal and on the activation feeding
into that weight. If the previous activation is doubled, the same small weight
change has twice the local effect.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
      node distance=1.15cm,
      box/.style={draw, rounded corners, align=center, minimum width=2.2cm,
        minimum height=0.8cm},
      arrow/.style={-{Latex[length=2.6mm]}, thick}
    ]
    \node[box, fill=green!8] (x) {$x$};
    \node[box, fill=blue!8, right=of x] (h1) {$h_1$};
    \node[box, fill=blue!8, right=of h1] (h2) {$h_2$};
    \node[box, fill=orange!12, right=of h2] (yhat) {$f_\theta(x)$};
    \node[box, fill=red!8, right=of yhat] (loss) {$\ell$};

    \draw[arrow] (x) -- (h1);
    \draw[arrow] (h1) -- (h2);
    \draw[arrow] (h2) -- (yhat);
    \draw[arrow] (yhat) -- (loss);

    \draw[arrow, dashed] (loss.south) to[out=-90,in=-90]
      node[midway, below=0.35cm, fill=white, inner sep=1pt]{gradients}
      (yhat.south);
    \draw[arrow, dashed] (yhat.south) to[out=-90,in=-90] (h2.south);
    \draw[arrow, dashed] (h2.south) to[out=-90,in=-90] (h1.south);
  \end{tikzpicture}
  \caption{The forward pass evaluates the network from left to right. The
  backward pass propagates derivatives from the loss back through the stored
  intermediate quantities.}
\end{figure}

\noindent
For ReLU hidden units, the backward recurrence includes the pointwise derivative
of the activation:
\begin{equation}
  \frac{\partial \ell_i}{\partial f_{k-1}}
  =
  \mathbb{I}[f_{k-1} > 0]
  \odot
  \left(
    \Omega_k^\top
    \frac{\partial \ell_i}{\partial f_k}
  \right),
\end{equation}
\noindent
where $\odot$ denotes elementwise multiplication. The indicator gates the
gradient: inactive ReLU units do not pass local gradient information backward.

The important point is not to memorize every indexed expression, but to know
what must be stored and what is reused. The activations from the forward pass
are needed during the backward pass. This makes backpropagation efficient, but
also memory-hungry for large networks.

\subsection{Automatic Differentiation and Practical Training}

Modern deep-learning libraries implement backpropagation through automatic
differentiation. A model is represented as an acyclic graph of operations. Each
operation knows how to compute its output in the forward direction and how to
propagate derivatives in the reverse direction. The user specifies the model,
the loss, and the optimizer; the framework records the graph and applies the
chain rule.

A typical training loop has the following structure:
\begin{enumerate}
  \item initialize the network parameters;
  \item choose a loss function and optimizer;
  \item load a batch of examples;
  \item run the forward pass and compute the loss;
  \item run the backward pass to compute gradients;
  \item update the parameters and repeat.
\end{enumerate}
\noindent
This is the conceptual core behind PyTorch-style code: zero old gradients,
evaluate the model, call backward on the loss, and let the optimizer step.

Several practical issues remain outside the pure gradient formula. Poor
initialization can make training slow or unstable. Too large a learning rate
can overshoot; too small a learning rate can make progress painfully slow.
Deep models may fit the training data while generalizing poorly. In the usual
bias--variance language, underfitting corresponds to high bias, while
overfitting corresponds to high variance. Regularization, more data, data
augmentation, dropout, weight decay, batch normalization, and early stopping
are common ways to improve generalization.

\begin{table}[ht]
  \centering
  \begin{tabular}{@{}p{3.0cm}p{5.2cm}p{5.0cm}@{}}
    \toprule
    Symptom & Meaning & Typical response \\
    \midrule
    High training error & The model or optimization setup is too weak &
      increase capacity, train longer, adjust learning rate \\
    Low training error but high validation error & The model overfits the
      training set & use regularization, more data, or early stopping \\
    Unstable loss & Updates are too aggressive or scales are poor &
      lower learning rate, improve initialization, normalize inputs \\
    \bottomrule
  \end{tabular}
  \caption{Most neural-network training failures can be diagnosed by comparing
  training behavior with validation behavior.}
\end{table}
